Bad forecast, good decision: predictive accuracy is not everything

### 2021-02-09 manuscript/appendix_B.Rmd
:pen:

The question of optimal harvest given a population growth model can be posed as a Markov Decision Process (MDP). @Marescot2013 provides an accessible overview of MDP problems and their solutions in conservation context.  @Reed1979 provides a formal proof of optimal harvest strategy for stochastic dynamics under the assumption of any concave growth function. The problem can be summarized as follows:

A manager seeks to maximize the sum of the utility derived from such a harvest
and such a state, $U(X_t,H_t)$, over all time, subject to discount rate $\delta$:

\begin{equation}
\sum_{t=0}^{t=\infty} U(X_t, H_t) \delta^t
\end{equation}

While in principle the utility could reflect many things, including the cost of fishing, market responses to supply and demand, the value of recreational fishing, the intrinsic value fish left in the sea [see @Halpern2013], for simplicity we will assume utility is merely a linear function of the harvest quota set by the manager, i.e. a fixed price $p$ per kilogram of fish harvested: $U(X_t, H_t) = p \min(H_t, X_t)$ (noting that realized harvest cannot exceed the available stock). As units are already specified in dimensionless quantities in this example, so without loss of generality we will set $p=1$.

To solve this optimization, we must also know the dynamics of $X_t$, describing how future stock $X_{t+1}$ changes from the present state, $X_t$, after to harvest $H_t$:

\begin{equation}
X_{t+1} = f(X_t, H_t)
\end{equation}

This problem is already well studied, and it is worth noting that even under the simple utility function which assigns no value to fish that are not harvested, the optimal strategy still seeks to sustain the fish population indefinitely; as @Clark1973 shows for the deterministic function $f$ and @Reed1979 extended to the stochastic case.  As @Reed1979 proves, the optimal strategy can be characterized by a policy of "constant escapement," in which the manager adjusts the harvest effort each year in an effort to ensure the same number or biomass of fish are left in the sea each year.

The optimization problem can then be solved by dynamic programming, using the recursive Bellman equation as ,

\begin{equation}
\end{equation}

This technique, known as stochastic dynamic programming (SDP) is widely used in conservation decision problems [e.g. @Mangel1985, @Marescot2013].

It is worth noting that while constant escapement is used in the management of several important fish stocks such as salmon, most marine fisheries are not managed using this optimal solution for a fluctuating population, but instead rely on a 'constant harvest effort' (or constant "yield") policy.  In the case of a deterministic model at equilibrium, these policies are identical.  Though constant-effort policies are technically sub-optimal for the problem as stated above with stochastic growth, in practice they can be more robust, such as when measurements of the current stock size $X_t$ are uncertain.  These issues are discussed in detail in @pomdp-intro.  For our purposes, it suffices to note that constant escapement policy is relevant both from a theoretical standpoint as the optimal solution for the problem under consideration, and an applied standpoint as a policy that guides management of many salmon fisheries.  The optimal control approach using SDP as illustrated here is used in a wide and growing number of conservation problems.

Bear in mind that in the example presented here, SDP merely serves the role of a convenient and established way for a manager to determine the optimal action, given some model estimation.  Whether we use SDP or some other way to determine the optimal policy, given the model, is immaterial to the conclusion.  The issue at hand is how we determine the model in the first place.

Given the function $f$ with known parameters, it is straight forward to determine the optimal harvest policy by stochastic dynamic programming (SDP, see @Mangel1985; @Marescot2013).

To solve the decision problem using SDP, we define the state space and the action space on a discrete grid of 240 points.  The maximum state is set well above the largest carrying capacity used in the models, which limits the influence of boundary effects introduced by the transformation to a discrete, finite grid.  Available harvest actions match the state space, effectively allowing any harvest level to be possible.

@@

The utility (reward) of an action is set to the (realized) harvest (e.g.
a fixed price per unit fish, with no cost applied to harvest effort).  Future
utility is discounted by fixed factor of $0.99$. Classic maximum sustainable 
yield models [@Schaefer1954] ignore discounting, while modern economic optimization
models insist on it, so a small discount conforms to the latter while reasonably
approximating the former.  The qualitative conclusion is not sensitive to the 
discounting rate.

@@

Alternate reward functions with varying price and cost structures are possible but do not qualitatively impact the conclusions.  This reward function is a limiting case of any more complex reward, and corresponds both to classic work that does not model utility explicitly [e.g. MSY theory, @Schaefer1954], as well as explicit assumptions typically made in more recent models [@Reed1979]. Moreover, using a simple reward function makes it clear that model that performs best is not doing so merely because of particular features baked into the a carefully chosen reward rule.

# Ecological Models

The manager chooses between two different logistic growth models, each of which
can be thought of as an approximation to the underlying "true" population model:

\begin{equation}
f_i(Y) = Y + Y r_i \left( 1 - \frac{Y}{K_i} \right) \xi_i(t)
\end{equation}

where $\xi_i(t)$ is a multiplicative log-normal noise term with mean 1 and log-standar-deviation $\sigma_i$.

Model 1 has $r_1 = 2$, $K = 16$, and $\sigma_1 = 0.05$.
Model 2 has $r_2 = 0.5, K_2 = 10$, and $\sigma_2 = 0.075$.

Meanwhile, the true population growth rate is simulated using a function with
non-linear per-capita growth:

\begin{equation}
f_i(Y) = Y + Y^4 r_i \left( 1 - \frac{Y}{K_i} \right)
\end{equation}

with $r_3 = 0.002$, $K_3 = 10$ and $\sigma_3 = 0.05$.

Here we provide annotated code necessary to completely reproduce all of the analysis presented in the main paper. This analysis is run in R [@R] uses `MDPtoolbox` [@MDPtoolbox] for solving Markov Decision Processes (MDP) using stochastic dynamic programming functionality, `expm` for matrix exponentials [@expm], and a few custom MDP functions provided by our package, `mdplearning` [@mdplearning].  We will also use `tidyverse`  packages for basic manipulation and plotting [@tidyverse]. This file is also available as an RMarkdown document [@rmarkdown] at  <https://github.com/cboettig/bad-forecast-good-decision>.

@@

## gather models together, indicate true model
sigma_g <- 0.05
models <- list("1" = f1, "2" = f2, "3" = f3)
model_sigmas <- c(sigma_g, 1.5 * sigma_g, sigma_g)
true_model <- "3"
```

@@

```{r}
transitions <- lapply(seq_along(models), 
                      function(i) transition_matrices(models[[i]], 
                                                      states, 
                                                      actions, 
                                                      model_sigmas[[i]]))
names(transitions) <- c("1", "2", "3")

@@

# Optimal control solutions

We use value iteration to solve the stochastic dynamic program [@Marescot2013; @MDPtoolbox] for each model. This determines the optimal harvest policy for each possible state, given each model. Because this step is the most computationally intensive routine, we cache the results using memosization conditioned on the transition matrices [@memoise].  Running this code with alternate transition matrices automatically invalidates that cache, reducing the risk of loading spurious results.

```{r sdp, results="hide"}
mdp <- memoise::memoise(mdp_value_iteration,
                        cache = memoise::cache_filesystem("cache/"))

policies <- 
  map_dfr(transitions, 
          function(P){
    soln <- mdp(P, reward, discount = discount, 
                epsilon = 0.01, max_iter = 2000, V0 = rep(0,dim(P)[[1]]))
    escapement <- states - actions[soln$policy]
    tibble(states, policy = soln$policy, escapement)
    }, 
  .id = "model")
```

@@

am1 <- adaptive_management(transitions[1:2], reward, discount, model_prior = c(0.99, 0.01), x0 = x0, 
               Tmax = 50, true_transition = transitions[[3]], 
                epsilon = 0.001, max_iter = 2000)
```

```{r}
am <- am1$df %>% mutate(belief = am1$posterior$V2)
am %>%  ggplot(aes(time, states[state], col = belief)) + 
  geom_line() + 
  geom_point() +
  scale_colour_gradient(limits = c(0,1), low = pal[1], high = pal[4])
```

@@

Using the transition matrices directly, we can examine what each model would 
have forecast the future stock size to be in the following year when no fishing
occurs (note that for each model, we use the transition matrix that corresponds
to 'no fishing', `model[[state_index, ,1]]`) (Fig 1a, main text).

The transition matrices give the full (discretized) probability distribution,
from which we can easily calculate both the expected value and the 95% 
confidence interval.

@@

We also look at the forecast each model makes when implementing the corresponding
optimal harvest:

@@

It is straight forward to apply the proper scoring formula of @Gneiting2007 
based on the first two moments of the distribution to score the respective
forecasts under both the unfished and actively managed scenario sfor each model:

@@

```{r plot_models}
d <- 
  map_dfc(models, function(f) f(states) - states) %>%
  mutate(state = states)

fig3a <- 
  d %>% pivot_longer(names(models), "model") %>%
  ggplot(aes(state, value, col=model, lty=model)) +
  geom_hline(aes(yintercept = 0), lwd=1) + 
  geom_line(lwd=2, show.legend = FALSE) + 
  coord_cartesian(ylim = c(-5, 8), xlim = c(0,16)) +
  ylab(bquote(f(x) - x)) + xlab("x")
```

>>>>

@@

@@

@@

n_models <- 32
rs <- seq(0.4, 2, length.out = n_models)
Ks <- seq(8, 10*8/5, length.out = n_models)
sigmas <- seq(0.05, 1.5*0.05, length.out = n_models)
pars <- data.frame(r = rs, K = Ks, sigma = sigmas)

closure <- function(r, K){
  function(x, h = 0){
  s <- pmax(x - h, 0)
  s + s * (r * (1 - s / K) )
  }
}

models <- lapply(1:n_models, function(i) closure(pars[i,"r"], pars[i,"K"]))

## gather models together, indicate true model
sigma_g <- 0.05
models <- c(f3, models)
model_sigmas <- c(sigma_g, pars$sigma)
true_model <- 1
```

@@

```{r}
transitions <- lapply(seq_along(models), 
                      function(i) transition_matrices(models[[i]], 
                                                      states, 
                                                      actions, 
                                                      model_sigmas[[i]]))
names(transitions) <- as.character(seq_along(models))

@@

# Plot Models

```{r plot_models}
model_set <- models
names(model_set) <- c("true", as.character(round(Ks, 2)))
d <- 
  map_dfc(model_set, function(f) f(states) - states) %>%
  mutate(state = states)

  d %>% pivot_longer(names(model_set[-1]), "model") %>%
  ggplot(aes(state, value, col=as.numeric(model))) +
  geom_hline(aes(yintercept = 0), lwd=1) + 
  geom_point(alpha = .5) + 
  geom_line(aes(state, true), col = "red") +
  coord_cartesian(ylim = c(-5, 8), xlim = c(0,16)) +
  ylab(bquote(f(x) - x)) + xlab("x") + viridis::scale_color_viridis()
```

@@

am1 <- adaptive_management(transitions[-1], 
                           reward, 
                           discount, 
                           x0 = x0, 
                           Tmax = 50, 
                           true_transition = transitions[[true_model]], 
                           epsilon = 0.001, 
                           max_iter = 2000)
```

```{r}
names(am1$posterior) <- Ks
belief <- am1$posterior %>%
  mutate(time = 1:dim(am1$posterior)[1]) %>% 
  pivot_longer(as.character(Ks), "K", values_to = "belief") %>%
  mutate(K = as.numeric(K))

belief %>% filter(time %in% seq(1, 11, by = 2)) %>%
  ggplot(aes(K, belief, col=time, group=as.character(time))) +
  geom_line()

belief %>% filter(K < 10) %>%
  ggplot(aes(time, belief, col=K, group=as.character(K))) +
  geom_line()

```{r}
am <- am1$df %>% mutate(belief = am1$posterior[,1])
am %>%  ggplot(aes(time, states[state], col = belief)) + 
  geom_line() + 
  geom_point() +
  scale_colour_gradient(limits = c(0,1), low = pal[1], high = pal[4])
```

@@

@@

@@

@@

### 2021-02-09 manuscript/manuscript.Rmd
:pen:

Global change issues are complex and outcomes are difficult to predict, making approaches to uncertainty a central part
of effective decision-making [@ @]. While some uncertainty is intrinsic to the underlying processes (stochasticity) or by
limits to what variables we can observe and how accurately we measure them (measurement uncertainty), much uncertainty 
is the result of our imperfect knowledge of the processes involved, as expressed by the structure and
parameters of mathematical models used to approximate those processes (model uncertainty). Model uncertainty can be 
reduced by gathering additional data, which may rule out certain models or parameter values as implausible. 
When model parameters are estimated directly from available data, there is a risk that the best-fitting models may 
_overfit_ to patterns arising from by chance from the stochasticity or measurement error which do not reflect the
underlying process. Researchers have often favored simpler models which are less prone to over-fitting, sometimes
through explicit penalties such as information criteria [@], though these can be misleading [@]. With the rapid 
expansion of available ecological and environmental data [@], it is increasingly possible to rely instead on comparisons
between model predictions and data purposely excluded from model estimation (cross-validation), or better, 
comparisons between model predictions and data collected in the future (forecasting, @Clark2001; @Dietze2018]). 
Over sufficiently long timescales, comparing model forecasts to future observations will select the models with
greatest predictive accuracy.  Because model-based forecasts frequently play an important role in decision-making,
it is commonly assumed [e.g. @Walters1981; @Clark2001; @Dietze2018] that this approach for addressing model uncertainty
will also improve decision-making outcomes. Here, I illustrate that this need not be the case: it is quiet possible
that any approach with improves the forecast accuracy of the model(s) over time can simultaneously lead to worse 
decision outcomes. This example illustrates that knowing when increased forecast accuracy will or will not also improve 
decision outcomes in general remains an open problem, as is the challenge of what to do about it.

@@

A wide range of paradigms are available for approaching the issue of decision-making under uncertainty.  These approaches can roughly be divided into two groups: the first group treats the issue of model uncertainty independently from the decision itself, while the second integrates the process of reducing model uncertainty into the process of decision making.  There are a wide range of techniques within each group.

@@

Like many decision problems, this one appears to hinge on having an accurate forecast: if we can predict to what size the stock will increase next year, $X_t+1$, and we know the current stock, $X_t$, then we can safely harvest $X_{t+1} - X_t$.  Overestimating or underestimating such recruitment will result in over-harvesting or under-harvesting, respectively.  Selecting a model based on forecast skill is also justifiable on theoretical grounds, since it reduces the risk of over-fitting by comparing model predictions to later observations that were not used to estimate the model. @Gneiting2007 provides a rigorous proof for "strictly proper" scoring criteria, which have the desirable property which no model predicting the distribution of future outcomes, $Q(x)$ can achieve a better average score than the true model $P(x)$ (That is, unlike likelihood or other goodness-of-fit scores, it is impossible to overfit when conditioning on a strictly proper score).

Iterative updating, with model priors.

>>>>

Global change issues are complex and outcomes are difficult to predict, making approaches to uncertainty a central part
of effective decision-making [@ @]. While some uncertainty is intrinsic to the underlying processes (stochasticity) or by
limits to what variables we can observe and how accurately we measure them (measurement uncertainty), much uncertainty 
is the result of our imperfect knowledge of the processes involved, as expressed by the structure and
parameters of mathematical models used to approximate those processes (model uncertainty). Unlike the other forms of
uncertainty, model uncertainty can be 
reduced by gathering additional data, which may rule out certain models or parameter values as implausible (). 
When model parameters are estimated directly from available data, there is a risk that the best-fitting models may 
_overfit_ to patterns arising from by chance from the stochasticity or measurement error which do not reflect the
underlying process. Researchers have often favored simpler models which are less prone to over-fitting, sometimes
through explicit penalties such as information criteria [@], though these can be misleading [@]. With the rapid 
expansion of available ecological and environmental data [@], it is increasingly possible to rely instead on comparisons
between model predictions and data purposely excluded from model estimation (cross-validation), or better, 
comparisons between model predictions and data collected in the future (forecasting, @Clark2001; @Dietze2018]). 
Over sufficiently long timescales, comparing model forecasts to future observations will select the models with
greatest predictive accuracy.  Because model-based forecasts frequently play an important role in decision-making,
it is commonly assumed [e.g. @Walters1981; @Clark2001; @Dietze2018] that this approach for addressing model uncertainty
will also improve decision-making outcomes. Here, I illustrate that this need not be the case: it is quiet possible
that any approach with improves the forecast accuracy of the model(s) over time can simultaneously lead to worse 
decision outcomes. This example illustrates that knowing when increased forecast accuracy will or will not also improve 
decision outcomes in general remains an open problem, as is the challenge of what to do about it.

@@

A wide range of paradigms are available for approaching the issue of decision-making under uncertainty. These approaches can roughly be divided into two groups: the first group treats the issue of model uncertainty independently from the decision itself, while the second integrates the process of reducing model uncertainty into the process of decision making to maximize the value of some objective. There are a wide range of techniques within each, and it is also possible to blend approaches. The key distinction is that methods in the first group do not involve any direct consideration of the possible actions or the utility that may result from those actions in how they select models [statistical approaches such as information criteria @BurnhamAnderson; and in particular, forecasting evaluation @Clark2001; @Dietze2018], while those in the second group require a more explicit statement of possible actions and the desired objectives [Decision theoretic approaches, for which @Polasky2011 provides an excellent and accessible review]. I illustrate how the most promising techniques from each of these groups would be applied to this simple problem, and demonstrate that it both cases they lead us away from the model that produces the most desirable decisions towards worse outcomes.  In retrospect, it will become obvious that neither these nor any other widely applied methods will select the model that leads to the best outcomes from the set of models considered.

@@

Like many decision problems, the task of setting a sustainable harvest quota appears to hinge on having an accurate forecast: if we can predict to what size the fish stock will increase next year, $X_t+1$, and we know the current stock, $X_t$, then we can safely harvest $X_{t+1} - X_t$.  Overestimating or underestimating such recruitment will result in over-harvesting or under-harvesting, respectively.  Selecting a model based on forecast skill is also justifiable on theoretical grounds, since it reduces the risk of over-fitting by comparing model predictions to later observations that were not used to estimate the model. @Gneiting2007 provides a rigorous proof for "strictly proper" scoring criteria, which have the desirable property which no model predicting the distribution of future outcomes, $Q(x)$ can achieve a better average score than the true model $P(x)$. That is, unlike likelihood or other goodness-of-fit scores, it is impossible to overfit when conditioning on a strictly proper score -- since no model model can out-perform the true model. Not that strictly proper scoring rules score _probabilistic forecasts_ and not just point predictions, favoring models which accurately reflect the uncertainty over those which under-estimate it.  These features have made proper scoring rules for probabilistic forecasts a successful and popular approach for addressing model uncertainty in many other areas [@Gneiting2014; @Raftery2016] and a promising tool for evaluation of ecological forecasts [@Dietze2018].

I illustrate the process of model selection by strictly proper scoring rules using two scenarios.

**Decision-Theoretic Approaches**

Decision-theoretic approaches include scenario analysis, resilience thinking, optimal control and related methods [@Polasky2011], which have a long history in ecology and particularly in fisheries management. One of the most influential of these is _adaptive management_ [sensu @Walters1978], in which a manager seeks to both reduce uncertainty over time while also achieving the best outcomes given current knowledge.  While the term today is frequently used in a looser sense, adaptive management as originally developed can be considered an example of optimal control under model uncertainty

[@Walters1981; @Smith1981]

# A way forward?

note that the model set considered here already contains a model which captures the key feature.

- Greedy strategy will not work.  Could stick with a model for 50 years then switch...
- A mechanistic understanding of the decision proces, not the ecological one.  Analyzing the decision problem, we know that in this situation, the optimal stategy depends only on the location of the peak of the growth rate.  This could suggest a very different way forward
-
